Distributional Regression Forests for Probabilistic Modeling and Forecasting

### 2018-02-09 pkg/disttree/inst/Psychoco2018/slides.Rnw
slides for Psychoco2018
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1998 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ffa37e05550d5c003b55bbc095743e245168dd3

\title{Distributional Trees and Forests}
\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
\URL{https://R-Forge.R-project.org/projects/partykit/}
\headerimage{3}

@@

\textbf{Distributional:} \\
\vspace{0.1 cm}
\hspace{0.5cm} Specify the complete probability distribution\\
\hspace{0.5cm} (including location, scale and shape).

@@

{\bf Forest:}\\
\begin{enumerate}
\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
\item Estimate $\hat{\theta}$ on the weighted learning data.
\end{enumerate}
% but weighted by similarities of the learning data and the new observation over all trees
%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
\end{frame}

\begin{frame}
\frametitle{Fitting a Tree}
\vspace*{-0.12cm}
\begin{center} 
DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
\vspace*{-0.21cm}
<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("x","x","x")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "   True parameters:   ",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "   True parameters:   ",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "   True parameters:   ",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)

@@

>>>>

\title{Distributional Regression Forests for Probabilistic Modeling and Forecasting}
\author{Achim Zeileis, Torsten Hothorn,\\ Lisa Schlosser, Heidi Seibold}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
\URL{https://R-Forge.R-project.org/projects/partykit/}
\headerimage{3}

@@

\textbf{Distributional:} \\
\vspace{0.1 cm}
\hspace{0.5cm} Specify the complete probability distribution\\
\hspace{0.5cm} (including location, scale, and shape).

@@

{\bf Forest:}\\
\begin{enumerate}
\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
\item Estimate $\hat{\theta}$ on the weighted learning data.
\end{enumerate}
% but weighted by similarities of the learning data and the new observation over all trees
%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Global likelihood estimation}
\begin{itemize}
\item Specified log-likelihood function $$\ell(\theta; y_i)$$ with distribution parameter $\theta$ and observation $y_i$ of a learning data set $\{y_i\}_{i = 1,\ldots,n}$.

\bigskip

\item Estimation of $\hat{\bold{\theta}}$ via maximum likelihood 
$$
\hat{\theta} = \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$
equivalent to solving 
$$
\sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
$$
\end{itemize}

\begin{frame}
\frametitle{Adaptive local likelihood estimation}
\medskip
{\bf Tree:}\\
\medskip
For a set of covariates $\bold{x}$, a learning data set $\{(y_i, \bold{x}_i)\}_{i =1,\ldots,n}$ and subgroups $\mathcal{B}_1, \ldots, \mathcal{B}_B$ calculate weights
$$
w^{\text{tree}}_i(\bold{x}) = \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b))
$$
which are then included in the estimation of $\hat{\theta}$
$$
\hat{\theta}_{\text{tree}}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{x}) \cdot \ell(\theta; y_i).
$$
\end{frame}

\begin{frame}
\frametitle{Adaptive local likelihood estimation}
\medskip
{\bf Forest:}\\
\medskip
For a set of covariates $\bold{x}$, a learning data set $\{(y_i, \bold{x}_i)\}_{i =1,\ldots,n}$, $T$~trees and corresponding subgroups $\mathcal{B}_1^t, \ldots, \mathcal{B}_B^t$ calculate weights
$$
w^{\text{forest}}_i(\bold{x}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
$$
which are then included in the estimation of $\hat{\theta}$
$$
\hat{\theta}_{\text{forest}}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{x}) \cdot \ell(\theta; y_i).
$$
\end{frame}

\begin{frame}
\frametitle{Adaptive local likelihood estimation}
{\bf Base model:}\\
\vspace{-0.5cm}
$$
\hat{\theta}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
$$

{\bf Tree:}\\
\vspace{-0.5cm}
$$
w^{\text{tree}}_i(\bold{x}) = \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b))
$$
$$
\hat{\theta}_{\text{tree}}(\bold{x}) = \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{tree}}_i(\bold{x}) \cdot \ell(\theta; y_i).
$$

{\bf Forest:}\\
\vspace{-0.5cm}
$$
w^{\text{forest}}_i(\bold{x}) = \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
$$
$$
\hat{\theta}_{\text{forest}}(\bold{x}) =  \max_{\theta \in \Theta} \sum_{i=1}^n w^{\text{forest}}_i(\bold{x}) \cdot \ell(\theta; y_i)
$$
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Fitting a Tree}
\vspace*{-0.12cm}
\begin{center} 
DGP:  $\; Y\ |\ X = x \; \sim  \; \mathcal{N}(\mu(x), \sigma^2(x))$
\vspace*{-0.21cm}
<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("x","x","x")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "   True parameters:   ",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "   True parameters:   ",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "   True parameters:   ",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)

@@

\begin{frame}
\frametitle{Choosing a distribution}

\medskip
{\bf Based on:}
\begin{itemize}
\item Recommendations by subject matter literature.
\item Experiences from other modelling situations.
\item Common simplifications/generalizations/assumptions.
\end{itemize}

\bigskip
{\bf Examples:}
\begin{itemize}
\item Survival times: Weibull distribution, exponential distribution.
\item Wages, precipitation: censored normal distribution.
\item Temperature: normal distribution.
\item Count of events/accidents (insurance): Poisson distribution.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Alternative: Transformation trees and forests}

{\bf General framework / Conditional model:}
$$
\mathbb{P}(Y\leq y|X=x) = \mathbb{P}_{Y|X=x} = \mathbb{P}_{Y, \theta(x)} 
$$

\bigskip

{\bf Transformation model:}
$$
\mathbb{P}(Y\leq y|X=x) = \Phi(\bold{a}_{Bs,d}(y)^T \theta(x)) 
$$

\begin{itemize} 
\item $\bold{a}_{Bs,d}(y)^T \theta(x)$ is a smooth, monotone Bernstein polynomial of degree $d$.
\item $d=1$ means $\mathbb{P}(Y\leq y|X=x) = \mathcal{N}(\mu(x),\sigma^2(x))$.
\item $d=5$ is surprisingly flexible.
\end{itemize}
%All ``classical'' distributions:  Distributional forests
\end{frame}
